Railway: Adaptive Disk Storage for Interaction Graphs

### 2014-09-11 docs/edbt2015/text/introduction.tex
some more progress

In our earlier work, we introduced an interaction graph database~\cite{gedik14}
that works on this principle of access locality. It proposed a disk
organization that consists of as set of blocks, each containing a list of
\emph{temporal neighbor lists}. A temporal neighborlist contains a head vertex
and a set of incident edges within a given time range. The layout optimizer
aims at bringing together temporal neighborlists that are (i) close in terms
of their temporal ranges, (ii) have many edges between them, and (iii) have
few edges going into temporal neighborlists outside the block.

% Attributes associated with interactions
% I/O inefficiency due to attributes that are not accessed
% Similar to the problem of row vs column stores

Attributes can be stored in two ways, either separately
(e.g., in a relational table), or locally with the temporal neighbor lists. 
If they are stored separately, then the graph database cannot take advantage
of locality optimizations performed for block organization.  As discussed in
Gedik et al.~\cite{gedik14}, the database must go back and forth between the
disk blocks to access the edge attributes.  On the other hand, if attributes
are stored locally in the disk blocks containing the graph structure, then
there can be significant overhead due to disk I/O if only a few attributes are
needed to answer a query.

% Challenges: finding the right partitioning
% what if the access pattern changes over time
% query access patterns are different over different time ranges

% Solution: railway layout 
% adaptive layout optimization
% algorithms for optimal partitioning
% heuristics & evaluation

Graph databases, and \emph{interaction graph} databases in particular, are a
critical component for many of today's most popular applications.  Graph
databases store entities as vertices, relationships as edges, and have data
associated with either vertices or edges, called \emph{attributes} or
\emph{properties}.  Interaction graphs differ from general graphs in that they
are \emph{append-only}, and edges are assigned a \emph{temporal value}.  Thus,
interaction graphs encode the relationships over time between entities in
their structure, making them ideal for use in social networking,
telecommunications, or modeling the world-wide web.

@@

It is widely recognized that query workload and disk layout have a significant
impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
table-based, relational databases, this fact has led database designers to
develop alternative approaches for storage layout: row-oriented storage is
more efficient when queries access many attributes from a small number of
records, and column-oriented storage is more efficient when queries access a
small number of attributes from many records~\cite{stonebraker05}. 
Unfortunately, although interaction graph databases, like relational
databases, are the target of diverse query workloads, there is no clear
correspondence to a row-oriented or column-oriented storage layout.

This paper presents an algorithm for adaptively optimizing disk block storage
for interaction graphs. The key idea behind the approach is a novel storage
scheme, called a \emph{rail layout}.  Initially, attributes are stored with
their neighbor lists in large blocks. Over time, the blocks are split into
smaller-blocks that contain the neighbor lists, but only subsets of the
attributes, depending on the query workload. Intuitively, this is almost like
having two or more graph databases for certain time regions, each containing a
different subset of the attributes, but with a link between them in case a
query needs to access both.

>>>>

In our earlier work~\cite{gedik14}, we introduced an interaction graph database
that works on this principle of access locality. It proposed a disk
organization that consists of as set of blocks, each containing a list of
\emph{temporal neighbor lists}. A temporal neighborlist contains a head vertex
and a set of incident edges within a given time range. The layout optimizer
aims at bringing together temporal neighborlists that are (i) close in terms
of their temporal ranges, (ii) have many edges between them, and (iii) have
few edges going into temporal neighborlists outside the block.

Many real world graph databases contain attributes. In the case of interaction
graphs, the attributes can be considered as properties associated with the
edges. Attributes can be stored in two ways, either separately (e.g., in a
relational table), or locally with the temporal neighbor lists.  If they are
stored separately, then the graph database cannot take advantage of locality
optimizations performed for block organization. The database must go back and
forth between the disk blocks to access the edge attributes.  On the other
hand, if attributes are stored locally in the disk blocks containing the graph
structure, then there can be significant overhead due to disk I/O if only a
few attributes are needed to answer a query.

@@

It is widely recognized that query workload and disk layout have a significant
impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
table-based, relational databases, this fact has led database designers to
develop alternative approaches for storage layout: row-oriented
storage~\cite{rowOrg} is more efficient when queries access many attributes
from a small number of records, and column-oriented storage~\cite{colOrg} is
more efficient when queries access a small number of attributes from many
records~\cite{stonebraker05}.  Unfortunately, although interaction graph
databases, like relational databases, are the target of diverse query
workloads, there is no clear correspondence to a row-oriented or
column-oriented storage layout.

This paper presents an adaptive disk layout called the \emph{railway layout}
and associated algorithms for optimizing disk block storage for interaction
graphs. The key idea is to divide blocks into one or more sub-blocks, where
each sub-block contains a subset of the attributes (potentially overlapping),
but the entire graph structure is replicated within each sub-block. This way,
a query can be answered completely by only reading the sub-blocks that contain
the attributes of interest, reducing the overall I/O.

There a number of challenges in achieving an effective adaptive layout. First,
we need to find the partitioning of attributes that optimizes the query I/O.
To address this, we model the problems of overlapping and non-overlapping
attribute partitioning as mixed-integer linear programs (ILPs), and provide
optimal solutions that minimize the I/O cost. Second, the query workload, and
thus the attribute access pattern can change over time. For this purpose, our
railway layout supports customization of the attribute partitioning of
sub-blocks on a per-block basis. Third, such flexibility necessitates online
configuration of attribute partitioning as the query workload evolves, which
in turn requires fast algorithms for performing the attribute partitioning.
For this purpose, we develop greedy heuristic algorithms for both overlapping
and non-overlapping partitioning scenarios. These algorithms can scale to
larger number of attributes, yet provide close to optimal query I/O
performance. Finally, the railroad layout tradeoffs storage space to gain
improved query I/O performance. The storage overhead is more pronounced for
the case of overlapping partitioning. To address this, we limit the amount of
storage  overhead that can be tolerated with an upper limit, and integrate
this limit to both our ILP formulations, as well as our greedy heuristics.

In summary, this paper makes the following contributions:
\begin{itemize}
\item We introduce the railroad layout for adaptive organization of interaction
graphs on disk. 
\item We introduce optimal ILP formulations for partitioning disk blocks into 
sub-blocks with overlapping and non-overlapping attributes, given a query
workload. Our formulation also support upper bounding the amount of storage
overhead introduced as a result of the railroad layout.
\item To support online adaptation, we develop greedy heuristics that can scale
better compared to the ILP alternatives, yet achieve close to optimal query
I/O.
\item We provide an extensive experimental study comparing our approach to a
few baseline alternatives.
\end{itemize}

The rest of the paper is organized as follows. Section...
